\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2022-05-10 manuscript/paper.tex
add theorem, update real-time prediction paradigme

\newtheorem{assumption}{Assumption}{\bf}{\rm} 
\newtheorem{lemma}{Lemma}

@@

Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
\[ \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\} \]
where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that
\[ y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \quad \text{ with } \quad y_{ij}^l=Y_i^l(t_{ij}^l) \]
for all $l=1, \ldots, L$. 
For the $i$-th subject, let us denote 
\[
\left\{
    \begin{array}{ll}
        y_i &= ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i},\\
        b_i &= ({b_i^1}^\top \cdots {b_i^L}^\top)^\top \in \R^r,
    \end{array}
\right.
\]
with $n_i = \sum_{l=1}^L n_i^l$ and $r = \sum_{l=1}^L r_l$ the total number of longitudinal measurements (for subject $i$) and the total dimension of the random effects respectively, as well as the following design matrices
\[ U_i = 
\begin{bmatrix}
  U_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & U_{iL}
\end{bmatrix} 
\in \R^{n_i \times q}
\qquad \text{and} \qquad
V_i = 
\begin{bmatrix}
  V_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & V_{iL}
\end{bmatrix}
\in \R^{n_i \times r}
\]
with $q = \sum_{l=1}^L q_l$ and where for all $l=1, \ldots, L$, one writes
\[
\left\{
    \begin{array}{ll}
        U_{il} &= \big(u_i^l(t_{i1}^l)^\top \cdots u_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times q_l},\\
        V_{il} &= \big(v_i^l(t_{i1}^l)^\top \cdots v_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times r_l}.
    \end{array}
\right.
\]
From now on, all computations are done conditionally on the design matrices $(U_i)_{i=1, \dots, n}$ and $(V_i)_{i=1, \dots, n}$.
Denoting $(t_1^u, \ldots, t_J^u)$ the $J \in \N_+$ unique failure times, we then extract all representation vectors $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}$ for all subjects $i=1, \dots, n$ before inference and from now on, all computations are also done conditionally on them.
Now, let us make two first hypothesis.
\begin{assumption}
\label{assumption1}
Suppose that the random effects are independent of the latent class membership, and that the latter remain independent conditional on the observed data (namely $T$, $\Delta$ and $Y$).
\end{assumption}
Assumption~\ref{assumption1} states that subject-and-longitudinal outcome specific random effects $b^l$ do not depend on the latent class membership, which is not a strong modeling assumption.
\begin{assumption}
\label{indep-hyp-2}
Suppose that for a given subject $i$ and conditional on the random effects and the latent class membership, all $Y_i^l$ are independent for $l=1, \dots, L$.
\end{assumption}
Assumption~\ref{indep-hyp-2} is similar to standard modeling hypothesis, see for instance~\citet{tsiatis2004joint}.
For all $k = 0, \ldots, K-1$, we denote 
\[ \beta_k = ({\beta_k^1}^\top \cdots {\beta_k^L}^\top)^\top \in \R^q \] 
and 
\[M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}. \]

@@

Let us denote 
\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
the collection of the $\vartheta \in \N_+$ unknown parameters to estimate. Note that we include $\lambda_0$ in $\theta$. Indeed, we will estimate the baseline using nonparametric maximum likelihood, as a function taking mass at each failure time. So in $\theta$, $\lambda_0$ represents the vector of these baseline hazard quantities, with a dimension equal to $J$.
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
\begin{assumption}
\label{indep-hyp-3}
Assume that
\begin{equation}
  \label{eq:ind-hyp}
  f_\theta(t_i, \delta_i, y_i | b_i, G_i = k) = f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | b_i, G_i = k)
\end{equation}
for all $i=1, \ldots, n$.
\end{assumption}
Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}. Remember that the densities are also conditionned by all the representation features $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}$. Hence Assumption~\ref{indep-hyp-3} says that the random effects vector $b_i$, the latent class membership and the representation features account for the association between the longitudinal and event outcomes, which makes sense given Figure~\ref{fig:graphical-model}.
Then, one has
\begin{equation}
	\label{eq:log-lik}
	\ell_n(\theta) = \ell_n(\theta ; \cD_n) = n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | G_i = k),
\end{equation}
where 
\[f_\theta(t_i, \delta_i| G_i = k) = \big[\lambda(t_i|\cY_i(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cY_i(s), G_i = k) \dd s \Big\}. \]
As expected, one has a JLCMs like log-likelihood.

@@

In this section, we describe the procedure for estimating the parameters of the lights model. Let us first present the considered penalized objective, and then focus on the algorithm proposed for inference.

@@

We need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters given by 
\begin{equation*}
  \cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].
\end{equation*}
Given Assumption~\ref{assumption1}, the previous expression requires to compute expectations of the form
\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \]
for different functions $g$, where we denote 
\begin{equation}
  \label{eq:pi_ik-def}
  \pi_{ik}^{\theta^{(w)}} = \P_{\theta^{(w)}}[G_i = k | t_i, \delta_i, y_i] 
\end{equation}
the posterior probability of the latent class membership using parameters $\theta^{(w)}$. 
Then, either $g(b_i, G_i) = \tilde g(b_i)$ (either $\tilde g_1: b_i \mapsto b_i$ or $\tilde g_2: b_i \mapsto b_i b_i^\top$), or $g(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.

@@

\paragraph*{Proximal Quasi-Newton M-step.}
Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
We then present the parameters updates in the order given in Algorithm~\ref{alg:prox-QNEM}.
First, the update of $D^{(w)}$ is naturally given in closed-form by
\begin{equation}
  \label{eq:update_D}
  D^{(w+1)} = n^{-1} \sum_{i=1}^n \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
\end{equation}
Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
\[ P^{(w)}_{n, k}(\xi_k) = -n^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-xi_k}
  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
\end{equation}
It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:pi_ik-def}). We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
of the inverse Hessian matrix of the objective function. It can deal with differentiable convex objectives with box constraints.
In order to use it with $\ell_1$ penalization, which is not differentiable, we use the trick borrowed from \citet{andrew2007scalable}: for $a \in \R$, write $|a| = a^+ + a^-$, where $a^+$ and $a^-$ are respectively the positive and negative part of $a$, and add the constraints $a^+ \geq 0$ and $a^- \geq 0$.
Namely, we rewrite the minimization problem~\eqref{eq:minimization-xi_k} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-xi_k}
  \begin{split}
    \text{minimize}& \quad \quad P_{n, k}^{(w)}(\xi_k^+ - \xi_k^-) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p (\xi_{k,j}^+ + \xi_{k,j}^-) + \dfrac \eta 2 \norm{{\xi_k}^+ - {\xi_k}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad \xi_{k,j}^+ \geq 0 \text{ and } \xi_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p
  \end{split} 
\end{equation}
where $\xi_k^\pm = (\xi_{k,1}^\pm, \ldots, \xi_{k,p}^\pm)^\top$. The L-BFGS-B solver requires the exact value of the gradient, which is easily given by
\begin{equation}
  \label{eq:grad-xi_k}
  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n  \big(\pi_{ik}^{\theta^{(w)}} - \pi_{\xi_k}(x_i) \big) x_i.
\end{equation}
In practice, we use the \texttt{Python} solver \texttt{fmin\_l\_bfgs\_b} from \texttt{scipy.optimize}~\citep{virtanen2020scipy}. Note that for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the overall algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it. To speed up the overall algorithm, we then decrease the L-BFGS-B tolerance with the iterations $w$.

@@

\paragraph*{The prox-QNEM algorithm.}
Algorithm~\ref{alg:prox-QNEM} describes the main steps of the resulting prox-QNEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}.

% \vspace{.5cm}

\begin{algorithm}[!htb]
\DontPrintSemicolon
\caption{prox-QNEM algorithm for the lights model inference}
\label{alg:prox-QNEM}
\KwInput{Training data $\cD_n$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})_{k=0,\ldots,K-1}$}
\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
\tcc{Initialization}
Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
\For{$w = 1,\ldots,$ until convergence}    
{ 
  \tcc{E-step}
  Compute $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:Lambda_ik-b_i} for $\tilde g_1$ and $\tilde g_2$\\
  Compute $\pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:pi_ik-def}\\
  \tcc{Proximal Quasi-Newton M-step}
  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:update_beta} \\
  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:gamma-update} \\
  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
}
\textbf{Return:} {$\hat \theta$}
\end{algorithm}

@@

@@

The hyper-parameters $(\zeta_1, \zeta_2)^\top \in \R_+^2$ are then tuned using a Bayesian optimization method~\citep{snoek2012practical}, with predictive marker defined in Section~\ref{sec:evaluation strategy} and the C-index score defined in Section~\ref{sec:Metrics}. In practice, we use the \texttt{Python} package \texttt{Hyperopt}~\citep{bergstra2013hyperopt} and the Tree-structured Parzen Estimator~\citep{bergstra2011algorithms}, which is a sequential model-based optimization (SMBO) approach. SMBO methods sequentially construct models to approximate the performance of hyperparameters based on historical measurements, and then subsequently choose new hyperparameters to test based on this model.
Moreover, the search interval for hyper-parameter $\zeta_1$ is automatically computed in a data driven way described in Appendix~\ref{sec:auto-grid-cv}, while the search intervals for $\zeta_2$ is determined empirically.
% TODO: Figure de visualisation de la CV de (\zeta_1, \zeta_2) en 3D

@@

In this section, we first present the proposed evaluation strategy to assess real-time prognostic prediction performances. We then introduce the appropriate metric used to this end: the C-index ; and finally, the considered models for performance comparisons.

\subsection{Evaluation strategy: the real-time prediction paradigm}
\label{sec:evaluation strategy}

When validating predictions or comparing performances of competing models in a survival context, one can be interested either in the discriminative power of the predictive rule and use concordance measures such as the C-index~\citep{heagerty2005survival} (defined in Section~\ref{sec:Metrics}), or in the predictive accuracy of the rule~\citep{schemper2000predictive}.
Developments in the field of joint modeling have primarily focused on modeling and estimation, and most studies do not consider goodness-of-fit nor generalization power in a prognostic prediction perspective~\citep{hickey2016joint}. However, with the prospect of using prognostic prediction in real-time on a daily basis, practitioners will naturally require predictive prognostic tools to evaluate models fit and compare models in terms of discriminative power. So we choose to put ourselves in what we call a ``real-time prediction paradigm''.

>>>>

\newtheorem{assumption}{Assumption}{\bf}{\rm} 
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}

@@

Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
\begin{equation}
  \label{eq:dataset-def}
  \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\}
\end{equation}
where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that
\[ y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \quad \text{ with } \quad y_{ij}^l=Y_i^l(t_{ij}^l) \]
for all $l=1, \ldots, L$. 
For the $i$-th subject, let us denote 
\[
\left\{
    \begin{array}{ll}
        y_i &= ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i},\\
        b_i &= ({b_i^1}^\top \cdots {b_i^L}^\top)^\top \in \R^r,
    \end{array}
\right.
\]
with $n_i = \sum_{l=1}^L n_i^l$ and $r = \sum_{l=1}^L r_l$ the total number of longitudinal measurements (for subject $i$) and the total dimension of the random effects respectively, as well as the following design matrices
\[ U_i = 
\begin{bmatrix}
  U_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & U_{iL}
\end{bmatrix} 
\in \R^{n_i \times q}
\qquad \text{and} \qquad
V_i = 
\begin{bmatrix}
  V_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & V_{iL}
\end{bmatrix}
\in \R^{n_i \times r}
\]
with $q = \sum_{l=1}^L q_l$ and where for all $l=1, \ldots, L$, one writes
\[
\left\{
    \begin{array}{ll}
        U_{il} &= \big(u_i^l(t_{i1}^l)^\top \cdots u_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times q_l},\\
        V_{il} &= \big(v_i^l(t_{i1}^l)^\top \cdots v_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times r_l}.
    \end{array}
\right.
\]
From now on, all computations are done conditionally on the design matrices $(U_i)_{i=1, \dots, n}$ and $(V_i)_{i=1, \dots, n}$.
Denoting $(t_1^u, \ldots, t_J^u)$ the $J \in \N_+$ unique failure times, we then extract all representation vectors $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}$ for all subjects $i=1, \dots, n$ before inference and from now on, all computations are also done conditionally on them.
Now, let us make two first hypothesis.
\begin{assumption}
\label{assumption1}
Suppose that the random effects are independent of the latent class membership, and that the latter remain independent conditional on the observed data (namely $T$, $\Delta$ and $Y$).
\end{assumption}
Assumption~\ref{assumption1} states that subject-and-longitudinal outcome specific random effects $b^l$ do not depend on the latent class membership, which is not a strong modeling assumption.
\begin{assumption}
\label{indep-hyp-2}
Suppose that for a given subject $i$ and conditional on the random effects and the latent class membership, all $Y_i^l$ are independent for $l=1, \dots, L$.
\end{assumption}
Assumption~\ref{indep-hyp-2} is similar to standard modeling hypothesis, see for instance~\citet{tsiatis2004joint}.
For all $k = 0, \ldots, K-1$, we denote 
\[ \beta_k = ({\beta_k^1}^\top \cdots {\beta_k^L}^\top)^\top \in \R^q \] 
and 
\[M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}. \]

@@

Let us denote 
\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
the collection of the $\vartheta \in \N_+$ unknown parameters to estimate. Note that we include $\lambda_0$ in $\theta$. Indeed, we will estimate the baseline using nonparametric maximum likelihood, as a function taking mass at each failure time. So in $\theta$, $\lambda_0$ represents the vector of these baseline hazard quantities, with a dimension equal to $J$.
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$ (defined in~\eqref{eq:dataset-def}), corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
\begin{assumption}
\label{indep-hyp-3}
Assume that
\begin{equation}
  \label{eq:ind-hyp}
  f_\theta(t_i, \delta_i, y_i | b_i, G_i = k) = f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | b_i, G_i = k)
\end{equation}
for all $i=1, \ldots, n$.
\end{assumption}
Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}. Remember that the densities are also conditionned by all the representation features $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}$. Hence Assumption~\ref{indep-hyp-3} says that the random effects vector $b_i$, the latent class membership and the representation features account for the association between the longitudinal and event outcomes, which makes sense given Figure~\ref{fig:graphical-model}.
Then, one has
\begin{equation}
	\label{eq:log-lik}
	\ell_n(\theta) = \ell_n(\theta ; \cD_n) = n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | G_i = k),
\end{equation}
where 
\[f_\theta(t_i, \delta_i| G_i = k) = \big[\lambda(t_i|\cY_i(t_i), G_i = k)\big]^{\delta_i} S_k(t_i) , \]
and
\begin{equation}
  \label{eq:survival-k-def}
  S_k(t_i) = \exp \Big\{-\int_0^{t_i} \lambda(s|\cY_i(s), G_i = k) \dd s \Big\}
\end{equation}
is the survival probability of subject $i$ given that it belongs to latent class $k$.
As expected,~\eqref{eq:log-lik} is a JLCMs like log-likelihood.

@@

In this section, we describe the procedure for estimating the parameters of the lights model. Let us first present the considered penalized objective, then focus on the algorithm proposed for inference, and on its convergence properties.

@@

We need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters given by 
\begin{equation}
  \label{eq:Q-def}
  \cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].
\end{equation}
Given Assumption~\ref{assumption1}, the previous expression requires to compute expectations of the form
\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \]
for different functions $g$, where we denote 
\begin{equation}
  \label{eq:pi_ik-def}
  \pi_{ik}^{\theta^{(w)}} = \P_{\theta^{(w)}}[G_i = k | t_i, \delta_i, y_i] 
\end{equation}
the posterior probability of the latent class membership using parameters $\theta^{(w)}$. 
Then, either $g(b_i, G_i) = \tilde g(b_i)$ (either $\tilde g_1: b_i \mapsto b_i$ or $\tilde g_2: b_i \mapsto b_i b_i^\top$), or $g(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.

@@

\paragraph*{Proximal Quasi-Newton M-step.}
Let us denote
\begin{equation}
  \label{eq:M-step-obj}
  \cQ_n^\text{pen}(\theta, \theta^{(w)}) = \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}
Here, we need to compute 
\[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n^\text{pen}(\theta, \theta^{(w)}).\]
We then present the parameters updates in the order given in Algorithm~\ref{alg:prox-QNEM}.
First, the update of $D^{(w)}$ is naturally given in closed-form by
\begin{equation}
  \label{eq:update_D}
  D^{(w+1)} = n^{-1} \sum_{i=1}^n \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
\end{equation}
Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
\[ P^{(w)}_{n, k}(\xi_k) = -n^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-xi_k}
  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
\end{equation}
It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:pi_ik-def}). We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
of the inverse Hessian matrix of the objective function. It can deal with differentiable convex objectives with box constraints.
In order to use it with $\ell_1$ penalization, which is not differentiable, we use the trick borrowed from \citet{andrew2007scalable}: for $a \in \R$, write $|a| = a^+ + a^-$, where $a^+$ and $a^-$ are respectively the positive and negative part of $a$, and add the constraints $a^+ \geq 0$ and $a^- \geq 0$.
Namely, we rewrite the minimization problem~\eqref{eq:minimization-xi_k} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-xi_k}
  \begin{split}
    \text{minimize}& \quad \quad P_{n, k}^{(w)}(\xi_k^+ - \xi_k^-) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p (\xi_{k,j}^+ + \xi_{k,j}^-) + \dfrac \eta 2 \norm{{\xi_k}^+ - {\xi_k}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad \xi_{k,j}^+ \geq 0 \text{ and } \xi_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p
  \end{split} 
\end{equation}
where $\xi_k^\pm = (\xi_{k,1}^\pm, \ldots, \xi_{k,p}^\pm)^\top$. The L-BFGS-B solver requires the exact value of the gradient, which is easily given by
\begin{equation}
  \label{eq:grad-xi_k}
  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n  \big(\pi_{ik}^{\theta^{(w)}} - \pi_{\xi_k}(x_i) \big) x_i.
\end{equation}
In practice, we use the \texttt{Python} solver \texttt{fmin\_l\_bfgs\_b} from \texttt{scipy.optimize}~\citep{virtanen2020scipy}. Note that for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the overall algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it. To speed up the overall algorithm, we then decrease the L-BFGS-B tolerance with the iterations $w$.

@@

\paragraph*{The prox-QNEM algorithm.}
Algorithm~\ref{alg:prox-QNEM} describes the main steps of the resulting prox-QNEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}. 
\begin{algorithm}[!htb]
\DontPrintSemicolon
\caption{prox-QNEM algorithm for the lights model inference}
\label{alg:prox-QNEM}
\KwInput{Training data $\cD_n$~\eqref{eq:dataset-def}; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})_{k=0,\ldots,K-1}$}
\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
\tcc{Initialization}
Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
\For{$w = 1,\ldots,$ until convergence}  
{ 
  \tcc{E-step}
  Compute $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:Lambda_ik-b_i} for $\tilde g_1$ and $\tilde g_2$\\
  Compute $\pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:pi_ik-def}\\
  \tcc{Proximal Quasi-Newton M-step}
  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:update_beta} \\
  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:gamma-update} \\
  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
}
\textbf{Return:} {$\hat \theta$}
\end{algorithm}

@@

\subsection{Convergence to a stationary point}

Let us address here convergence properties of the prox-QNEM algorithm described in the previous section. Since in~\eqref{eq:M-step-obj} we only minimize $\cQ_n^\text{pen}(\theta, \theta^{(w)})$ and not directly $\cQ_n(\theta, \theta^{(w)})$ defined in~\eqref{eq:Q-def} as in a classical EM~\citep{wu1983convergence}, we are in a generalized EM (GEM) setting~\citep{dempster1977maximum}. For such algorithm, one has the descent property saying that the objective function~\eqref{eq:pen-log-lik} decreases at each iteration, namely 
\[\ell^{\text{pen}}_n(\theta^{(w+1)}) \leq \ell^{\text{pen}}_n(\theta^{(w)}),\]
which can be observed in Figure ?.
Now, let us make a classical hypothesis.
\begin{assumption}
  \label{hyp:bounded}
  Suppose that $\ell^{\text{pen}}_n$ is bounded for all $\theta$.
\end{assumption}
Assumption~\ref{hyp:bounded} implies that $w \mapsto \ell^{\text{pen}}_n(\theta^{(w)})$ decreases monotically to a finite limit. Then, one has the following convexity property with the Gaussian case used in practice.
\begin{lemma}
\label{lem:strictly-convex}
$\cQ_n^{\text{pen}}(\theta, \theta^{(w)})$ is continuous in $\theta$ and $\theta^{(w)}$, and for any fixed $\theta^{(w)}$, it is a convex function in $\theta$, and strictly convex in each coordinate of $\theta$.
\end{lemma}
Hence, one can state the following theorem about the convergence of the prox-QNEM algorithm to a stationary point, being here a local minimum.
\begin{theorem}
\label{th:cvg}
Under Assumption~\ref{hyp:bounded}, every cluster point $\bar{\theta}$ of the sequence $\{ \theta^{(w)}\}_{w \geq 1}$ generated by the prox-QNEM algorithm is a stationary point of the objective function~\eqref{eq:pen-log-lik}, that is
\[\forall \varepsilon > 0, V_\varepsilon(\bar{\theta}) \cap \{ \theta^{(w)}\}_{w \geq 1} \setminus \{\bar{\theta}\} \neq \varnothing \implies \forall \tilde \theta \in \R^\vartheta, \underset{u \rightarrow 0}{\lim} \frac{\ell^{\text{pen}}_n(\bar{\theta} + \tilde \theta u) - \ell^{\text{pen}}_n(\bar{\theta})}{u} \geq 0 ,\]
with $V_\varepsilon(\bar{\theta})$ the epsilon-neighbourhood of $\bar{\theta}$.
\end{theorem}
The proof of Lemma~\ref{lem:strictly-convex} is straightforward, and we follow the proof of Theorem 1 in~\citet{bussy2019c} to prove Theorem~\ref{th:cvg}.

@@

The hyper-parameters $(\zeta_1, \zeta_2)^\top \in \R_+^2$ are then tuned using a Bayesian optimization method~\citep{snoek2012practical}, with the C-index score and the predictive marker both defined in Section~\ref{sec:evaluation strategy}. In practice, we use the \texttt{Python} package \texttt{hyperopt}~\citep{bergstra2013hyperopt} and the Tree-structured Parzen Estimator~\citep{bergstra2011algorithms}, which is a sequential model-based optimization (SMBO) approach. SMBO methods sequentially construct models to approximate the performance of hyperparameters based on historical measurements, and then subsequently choose new hyperparameters to test based on this model.
Moreover, the search interval for hyper-parameter $\zeta_1$ is automatically computed in a data driven way described in Appendix~\ref{sec:auto-grid-cv}, while the search intervals for $\zeta_2$ is determined empirically.
% TODO: Figure de visualisation de la CV de (\zeta_1, \zeta_2) en 3D

@@

In this section, we first present the proposed evaluation strategy to assess real-time prognostic prediction performances. We then introduce the considered models for performance comparisons.

\subsection{Evaluation strategy}
\label{sec:evaluation strategy}

When validating predictions or comparing performances of competing models in a survival context, one can be interested either in the discriminative power of the predictive rule and use concordance measures such as the C-index~\citep{heagerty2005survival} (defined in Section~\ref{sec:evaluation strategy}), or in the predictive accuracy of the rule~\citep{schemper2000predictive}.
Developments in the field of joint modeling have primarily focused on modeling and estimation, and most studies do not consider goodness-of-fit nor generalization power in a prognostic prediction perspective~\citep{hickey2016joint}. However, with the prospect of using prognostic prediction in real-time on a daily basis, practitioners will naturally require predictive prognostic tools to evaluate models fit and compare models in terms of discriminative power. So we choose to put ourselves in what we call a ``real-time prediction paradigm''.
